\protect A Web of Human Knowledge: Unravelling Wikipedia's First Link Network

### 2015-11-16 writeup/2015-11-wikipedia-network.tex
paper-esque version

\documentclass[twoside]{article}

% ------
% Fonts and typesetting settings
\usepackage[sc]{mathpazo}
\usepackage[T1]{fontenc}
\linespread{1.05} % Palatino needs more space between lines
\usepackage{microtype}
\usepackage{textcomp}

% ------
% Page layout
\usepackage[hmarginratio=1:1,top=32mm,columnsep=20pt]{geometry}
\usepackage[font=it]{caption}
\usepackage{multicol, hyperref}
\usepackage{parskip}

% ------
% Lettrines
\usepackage{lettrine}
\usepackage{color}

% ------
% Abstract
\usepackage{abstract}
	\renewcommand{\abstractnamefont}{\normalfont\bfseries}
	\renewcommand{\abstracttextfont}{\normalfont\small\itshape}

% ------
% Titling (section/subsection)
\usepackage{titlesec}
\renewcommand\thesection{\Roman{section}}
\titleformat{\section}[block]{\large\scshape\centering}{\thesection.}{1em}{}

% ------
% Header/footer
\usepackage{fancyhdr}
	\pagestyle{fancy}
	\fancyhead{}
	\fancyfoot{}
	\fancyhead[C]{ $\bullet$ Draft $\bullet$}
	\fancyfoot[RO,LE]{\thepage}

% My Shortcuts

@@

%shorcuts with arguments
\newcommand{\abs}[1]{\left\vert#1\right\vert} %nice absolute values
\newcommand{\bt}[1]{\textbf{#1}} %bold
\newcommand{\eq}[1]{\begin{align*}#1\end{align*}} %aligned equations
\newcommand{\cb}[1]{\centerline{\fbox{#1}}} %centered box
\newcommand{\bp}[1]{\fbox{\parbox{0.8\textwidth}{#1}}} %box paragraph
\newcommand{\norm}[1]{\left\lVert#1\right\rVert} %vector norm
\newcommand{\notimplies}{% does not imply
  \mathrel{{\ooalign{\hidewidth$\not\phantom{=}$\hidewidth\cr$\implies$}}}}
\renewcommand{\eq}[1]{\begin{align*}#1\end{align*}} %aligned equations

@@

\newcommand{\blue}[1]{{\leavevmode\color{lightblue}{#1}}} %solarized blue 
\newcommand{\green}[1]{{\leavevmode\color{javagreen}{#1}}} %command for green
\newcommand{\red}[1]{{\leavevmode\color{sred}{#1}}} %solarized red
\newcommand{\gray}[1]{{\leavevmode\color[gray]{0.5}{#1}}} %gray text

%environment
\newcommand{\tab}{\phantom{ssss}}
%----------------

% ------
% Clickable URLs (optional)
\usepackage{hyperref}

% ------
% Maketitle metadata
\title{\vspace{-5mm}%
	\fontsize{24pt}{12pt}\selectfont
	\textbf{What's so Special about Philosophy?} 
	}	
\author{%
\fontsize{14pt}{14pt}\selectfont
	Unraveling Wikipedia's First Link Network \vspace{-2mm}\\
	}
\date{}

%figures
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{float}

%text, figure spacing
\raggedbottom

%footnote withou marker
\newcommand\blfootnote[1]{%
  \begingroup
    \renewcommand\thefootnote{}\footnote{#1}%
      \addtocounter{footnote}{-1}%
   \endgroup }
%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\maketitle
\thispagestyle{fancy}
%========================ABSTRACT====================================

\begin{abstract}
\fontsize{12pt}{12pt}
\selectfont

Apples, oranges, and the most obscure Dylan song too---is everything a few clicks from Philosophy? 
Within Wikipedia, the surprising answer is yes: nearly all 
paths lead to Philosophy.
\footnote{
nearly $99.6\%$ of First Link paths, ending inside Wikipedia, lead to the Philosophy article (see traversal funnels discussion for details).}
\footnote{
An analysis performed in 2011 by Mat Kelcey posted on his blog (and a smiliar post by Ilmari Karonen) suggest $94.5\%$ of articles end up at Philosophy.
}
Wikipedia is the largest, most meticulously indexed collection of human knowledge ever amassed. 
\footnote{
cite MIT Tech Review Article: http://www.technologyreview.com/featuredstory/520446/the-decline-of-wikipedia/
}
More than information about a topic though, Wikipedia is a marvelous web of naturally emerging relationships.  
By following the First Link in an article, we connect entries to form a directed network within Wikipedia: Wikipedia's First Link Network. 
Here we study the English edition of Wikipedia's First Link Network for insight into how we relate topics, ideas, people, objects, and events.

We algorithmically parse all 4.7 million articles to construct a map of Wikipedia's First Link Network. 
In a novel approach to uncover structure, we traverse every possible path through the network, 
measuring the accumulation of First Links, path lengths, basins, cycles, and even particular articles funneling links into the cycles.
We discover many scale-free distributions, find Philosophy at a salient center, and uncover a flow from specific to general with 
basins around fundamental notions such as Community, State, and Science. 
Curiously, we also observe a gravitation towards topical articles including Health Care and Fossil Fuel.
These findings enrich our view of how we connect and structure
an ever growing load of information.

@@

\fontsize{11pt}{11pt}
\selectfont

%========================INTRODUCTION====================================
At no point in history has a larger or more meticulously indexed collection of human knowledge existed.
In amassing such an awe-inspiring collection, we formed an equally impressive web. 
Through the efforts of millions of individuals, working independently,
((cite))
we naturally linked topics, inventions, people, objects, places, and events across space and time.
The web we weaved, and continue to weave, is a wealth of information not only about those notable inventions, 
places, figures, and ideas, but also about \textit{relationships} among them.
We study the relationships in this naturally arising web through the hyperlinks connecting one article to another.

We build Wikipedia's First Link Network by following the First Link, not in parenthesis, inside the main body of each article in the English edition of Wikipedia. 
For the directed network to meaningfully reflect how we associate one article to another, we exclude links in parenthesis, 
disregarding pronunciation keys or disambiguations.
We also exclude any links in the side-bar elements, as well as any links to external pages, files, or WikiMedia 
projects outside of Wikipedia (such as Wiktionary).
This procedure corresponds to the original claim in 2008 about the percentage of pages with First Links to 
"Philosophy."\footnote{
https://en.wikipedia.org/wiki/Wikipedia:GettingtoPhilosophy
}
\footnote{
http://xkcd.com/903/
}
\footnote{
https://www.reddit.com/r/pics/comments/gpdhb/trythiswikipediamindfk/
}

The result is a directed network placing every article in a broader web of ideas.

\begin{figure}[H]
\centering
\caption{Wikipedia Train}
    \begin{subfigure}[b]{0.8\textwidth}
        \includegraphics[width=\textwidth]{graphics/wiki_train.png}    
    \end{subfigure}
\end{figure}

%===========================Data_and_Parsing================================
\section{Constructing the First Link Network}

Wikipedia renders and stores articles in MediaWiki markup, a markup langauge with syntax and keywords to format and mark elements in a page. Along with special syntax for links, MediaWiki markup includes templates for audio files, images, and side-bar
information.
While a human can accurately identify the First Link, to map the entire First Link Network of 11 million articles, we programmatically untangled the body text from side-bar, header box, and invalid link elements.

While some libraries exist for MediaWiki Markup, we opted to develop our own algorithm for parsing the First Link in the XML version of each article.\footnote{
Approaches using existing libraries led to several bugs 
including trouble with nested links, nested parenthesis, unclosed tags, escape characters 
as well as compatibility with other libraries used to parse the XML.}
Our parsing algorithm aimed to: 
1) accurately identify the First Link among other page elements
2) efficiently do so, without the need for several passes through the data.
To process an article with one pass, we developed a hierarchical system of flags:

\begin{figure}[H]
\centering
\caption{parsing algorithm of Wikipedia's XML dump}
    \begin{subfigure}[b]{0.5\textwidth}
        \includegraphics[width=\textwidth]{graphics/flags.pdf}    
        \caption{ The highest flag in the hierarchy indicates a Wikimedia template used to mark an element in the side bar, display an image, link to an external file, or another Wikimedia project outside of Wikipeida. Next, to catch any remaining elements outside the main body we have a second flag for <ref>, <div> elements. Finally, we catch parenthesis to ensure we do not capture a link to a pronunciation key.}
    \end{subfigure}

\end{figure}

To process the entirety of Wikipedia, we distributed the parsing and processing of the XML dump
across 112 cores of the UVM supercomputer cluster
((cite))
We then joined the results to form a hash table containing every Wikipedia article and its corresponding
First Link. The resulting network map is the basis of our analysis.

%========================Traversing the Network====================================
\section{Traversing the Network}

To understand the structure of Wikipedia's First Link Network, we
aimed to characterize the dynamics of the flow from one article to another. 
Do links tend towards a particular article, group of articles, or various groups of articles? 
What is the flow of links through a typical article?
What types of cycle (loop) structures exist in the network?
What are typical paths from one article to loop or an invalid link? 
Are there exceptionally long or short paths? 
Are there articles funneling the flow of links towards a particular path?
In answering these questions about the directed network, we aimed to characterize 
how so many independent articles relate to each other.

The n-degree, or number of links directly pointing to a particular article,
while a natural measure, fails to fully capture the dynamics needed to answer these questions.
The n-degree measures only the particular set of links to a particular article, 
rather than the richer dynamics of how the links flow through the network: 
where articles tend, what the typical and atypical 
paths through the network are, which articles funnel relatively more links and so on.

Consequently, to capture the dynamics of how the First Links flow, we actually traverse every possible path through the network. 
We developed three metrics to uncover structure: traversal visits, traversal funnels, and path length. Each metric captures one essence of how the links flow. Traversal visits gauge the accumulation of links; traversal funnels gauge the influence on link path; path length measures the number of First Links to a repeated or invalid link.

The algorithm for traversing the network begins by selecting any article. We then proceed to the next article by following the First Link---recalling a First Link is a link in the main body of the article leading to another Wikipedia article. We repeat until the First Link is invalid or repeated to form a path. The collection of articles is path-connected conveying a flow of concepts from one article to another. 
The method is order agnostic with respect to which articles are selected first.
As long as each article is selected eventually, the resulting metrics are equivalent.

\begin{figure}[H]
\centering
    \caption{traversal visit algorithm on a sample network}
\begin{subfigure}[b]{\textwidth}
    \includegraphics[width=\textwidth]{graphics/traversal_visit_algo_figure.pdf}
    \caption{The traversal visit vectors are an adjacency matrix for the paths through the network: the first column is indicates the path formed starting with article A. The number of traversal visits for article A is then the number of paths containing A or the sum of the first row in our matrix.}
\end{subfigure}
\end{figure}

The first metric, traversal visits, gauges where the flow of first accumulates by associating a count to each article. Every path through the article increments the associated count by 1. We can construct a vector to represent the articles incremented on a particular path, combining these vectors to form a matrix. The total accumulated counts for a given article is then simply the sum of the row in our matrix corresponding to the article of interest. Node A in our example has 7 traversal visits, since every path includes article A. Whereas, node E only has two traversal visits since only two paths contain article E for example.
What we have then is a metric signaling which article contains a larger accumulation of first links.

While traversal visits measure accumulation, we cannot gauge whether a particular article exerts a greater influence on flow of the first link paths. To distinguish between an article that simply happened to fall within a cycle from an article actually funneling the flow of first links, we developed a second metric called traversal funnels. We traverse the network in the same manner, but end our paths once we enter a cycle. We again increment the associated count of each article by one if the path up to a cycle contains the article. We call the accumulated count the number of traversal funnels. By measuring traversal funnels, we quantify
which articles funnel more link paths in a particular direction or cycle.

\begin{figure}[H]
\centering
    \caption{traversal funnel algorithm on a sample network}
\begin{subfigure}[b]{\textwidth}
    \includegraphics[width=\textwidth]{graphics/traversal_funnel_algo_figure.pdf}
    \caption{The algorithm for traversal funnels is identical to the previous algorithm for traversal visits with one alteration: the path ends at the start of a cycle to distinguish articles directing a path into a cycle from articles that simply happen to be in a highly traversed path. We can construct similar vectors by considering each path through the network, measuring traversal funnels for a particular article as the sum of the entries in its corresponding row.}
\end{subfigure}
\end{figure}

Finally, we track the number of links until a repeated article or invalid link, to find the typical path length within the network.
By recording the history of articles we traversed, we are also able to uncover several other network charactersitics including basins, defined as a collection of path-connected articles ranked by traversal visits, and cycles of various lengths.
The three metrics: traversal visits, traversals funnels, and path length, along with our path history
yield a powerful arsenal of information with which to study the network.

%========================Results====================================
\section{Discoveries}

@@

\begin{figure}[H]
\centering
    \caption{distribution of traversal visits}
        \includegraphics[width=\textwidth]{graphics/traversals_per_article.png}
\end{figure}

To more accurately gauge the distribution, we constructed a log-log plot of the entire dataset: log(traversal visits) against log(rank). We observe a fairly linear fit, as is characteristic of scale-free networks, with an r-value of -.93. A handful of the highest ranking articles contain a disproportionate number of traversal visits, while most have none. The skew in the distribution is not terribly surprising when considering the heuristic of how the links flow: from specific to general.

\begin{figure}[H]
\centering
\caption{traversal visit distribution of highest ranking articles}
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{graphics/top_1k_article_traversals.png}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{graphics/top_100_article_traversals.png}
\end{subfigure}
\end{figure}

@@

\begin{figure}[H]
\centering
    \caption{highest ranking articles by number of traversal visits}
        \includegraphics[width=0.6\textwidth]{graphics/articles_ranked.png}
\end{figure}

@@

\begin{figure}[H]
\centering
\caption{highest ranking 2-cycles}
    \begin{subfigure}[b]{0.8\textwidth}
        \includegraphics[width=\textwidth]{graphics/2_cycles.pdf}
    \end{subfigure}
\end{figure}

@@

\begin{figure}[H]
\centering
\caption{highest ranking 3-cycles}
    \begin{subfigure}[b]{0.8\textwidth}
        \includegraphics[width=\textwidth]{graphics/3_cycles.pdf}
    \end{subfigure}
\end{figure}

@@

\subsection{Basins}
We can group articles lying on the same path to identify basins of path-connected articles, not necessarily forming a perfectly closed cycle.
Ranking basins by traversal visits, we find many of highest ranking basins are around "Philosophy" as we might expect. 
Looking beyond "Philosophy" however, we find high ranking basins around similarly foundational ideas:
"Community", "Landmass", "Federal Government", "Presentation", and "Belief System". 
These concepts naturally emerge from the First Link Network potentially indicating pillars, which 
anchor specific knowledge in a broader, simpler concept.

\begin{figure}[H]
\centering
\caption{highest ranking 3-Cycles}
    \begin{subfigure}[b]{0.8\textwidth}
        \includegraphics[width=\textwidth]{graphics/basins.pdf}
    \end{subfigure}
\end{figure}

@@

\begin{figure}[H]
\centering
    \caption{Path Length Distribution}
    \begin{subfigure}[b]{0.5\textwidth}
        \includegraphics[width=\textwidth]{graphics/path_lengths_boxplot.png}
    \end{subfigure}
\end{figure}
\begin{figure}[H]
\centering
    \caption{Longest Path Lengths}
    \begin{subfigure}[b]{0.5\textwidth}
        \includegraphics[width=\textwidth]{graphics/top_1k_path_length.png}
    \end{subfigure}
\end{figure}

@@

Measuring traversal funnels reveals a dramatically different structure where "Philosophy" stands unmatched by orders of magnitude:

\begin{figure}[H]
\centering
    \caption{Top Funnels}
    \begin{subfigure}[b]{0.9\textwidth}
        \includegraphics[width=\textwidth]{graphics/funnels.pdf}
    \end{subfigure}
\end{figure}

@@

%========================Conclusion====================================
\section{Reflections}

@@

\blfootnote{
- thank you to my friend RJ for pointing out the xkcdc comic about links to Philosophy
}

@@

\section{Supplementary}

\begin{figure}[H]
\centering
\caption{highest ranking 2-Cycles}
    \begin{subfigure}[b]{0.8\textwidth}
        \includegraphics[width=\textwidth]{graphics/top_2loops.png}
    \end{subfigure}
\end{figure}

\begin{figure}[H]
\centering
\caption{highest ranking 2-Cycles}
    \begin{subfigure}[b]{0.8\textwidth}
        \includegraphics[width=\textwidth]{graphics/top_3loops.png}
    \end{subfigure}
\end{figure}
\begin{figure}[H]
\centering
    \caption{Top Funnels}
    \begin{subfigure}[b]{0.5\textwidth}
        \includegraphics[width=\textwidth]{graphics/top_funnels.png}
    \end{subfigure}
\end{figure}
%========================Style Guide====================================

%\newpage
%
%\section*{Style Guide}
%
%\begin{itemize}
%\item Web Terms
%    \begin{itemize}
%        \item \bt{article} instead of page
%        \item \bt{invalid link} (instead of dead or bad link)
%    \end{itemize}
%\item Measurements
%    \begin{itemize}
%        \item \bt{traversals visits} = node weight based on number of paths crossing an article.
%        \item \bt{path length} = number of articles traversed until an invalid link or a repeated article
%        \item \bt{traversal funnels} = traversals before a cycle is reached.
%    \end{itemize}
%\item Network Properties
%    \begin{itemize}
%        \item \bt{cycle} = permutation or a true loop
%        \item \bt{path-connected articles} for both cycles and articles along the same path leading to an 
%    invalid link
%        \item \bt{basin} a path-connected group of articles (not necessarily forming a cycle)
%    \end{itemize}
%\end{itemize}
%
%- add quotes around article names: Philosophy should be "Philosophy"
%
%- add thank you to RJ for pointing out the Blog Post

>>>>

\documentclass[pre,twocolumn,twoside,superscriptaddress,floatfix, aps, 10pt]{revtex4-1}

\usepackage{currfile}
\lefthyphenmin=3
\righthyphenmin=2

\usepackage{graphicx,epsfig,verbatim,enumerate}
\usepackage{amssymb,amsmath}
\usepackage{ifthen}

% Fonts and typesetting settings
\usepackage[sc]{mathpazo}
\usepackage[T1]{fontenc}
\linespread{1.05} % Palatino needs more space between lines
\usepackage{microtype}
\usepackage{textcomp}

\usepackage{longtable}

\usepackage{mathtools}

\newboolean{twocolswitch}

\newcommand{\www}[1]{\url{#1}}
\newcommand{\req}[1]{(\ref{#1})}
\newcommand{\Req}[1]{Eq.~(\ref{#1})}

% Lettrines
\usepackage{lettrine}

@@

%shorcuts with arguments
\newcommand{\abs}[1]{\left\vert#1\right\vert} %nice absolute values
\newcommand{\bt}[1]{\textbf{#1}} %bold
\newcommand{\eq}[1]{\begin{align*}#1\end{align*}} %aligned equations
\newcommand{\norm}[1]{\left\lVert#1\right\rVert} %vector norm
\renewcommand{\eq}[1]{\begin{align*}#1\end{align*}} %aligned equations

%piecewise function

%\begin{displaymath}
%   f(x) = \left\{
%     \begin{array}{lr}
%       1 & : x \in \mathbb{Q}\\
%       0 & : x \notin \mathbb{Q}
%     \end{array}
%   \right.
%\end{displaymath}

%environment
\newcommand{\tab}{\phantom{ssss}}

\usepackage{color}
\newcommand{\todo}[1]{\noindent\textcolor{blue}{{$\Box$ #1}}}

@@

\newcommand{\blue}[1]{{\leavevmode\color{lightblue}{#1}}} %solarized blue 
\newcommand{\green}[1]{{\leavevmode\color{javagreen}{#1}}} %command for green
\newcommand{\red}[1]{{\leavevmode\color{sred}{#1}}} %solarized red
\newcommand{\gray}[1]{{\leavevmode\color[gray]{0.5}{#1}}} %gray text
\newcommand{\Prob}[1]{{\rm Pr}\left(#1\right)}

\setboolean{twocolswitch}{true}

\begin{document}

\title{\protect
A Web of Human Knowledge: 
Unravelling Wikipedia's First Link Network
}

\author{
\firstname{Mark}
\surname{Ibrahim}
}
\email{mark.s.ibrahim@uvm.edu}

\affiliation{Department of Mathematics \& Statistics, 
    Computational Story Lab,  \\
    The University of Vermont, Burlington, VT 05401.}

\author{
\firstname{Christopher}
\surname{M. Danforth}
}
\email{chris.danforth@uvm.edu}

\affiliation{Department of Mathematics \& Statistics, 
    Computational Story Lab,  \\
    The University of Vermont, Burlington, VT 05401.}

\author{
\firstname{Peter}
\surname{Sheridan Dodds}
}

\email{peter.dodds@uvm.edu}

\affiliation{Department of Mathematics \& Statistics, 
    Computational Story Lab,  \\
    The University of Vermont, Burlington, VT 05401.}

\begin{abstract}
  \protect
  this is the glorious abstract of abstractness summarizing methods,
  results, and contribution of paper our understanding of the world

@@

\maketitle

\section*{Introduction}

Wikipedia is a towering achievement of the modern era. 
At no point in history has a larger or more meticulously indexed collection of human knowledge 
existed.
Wikipedia contains 37 million articles in 283 languages, 
with coverage spanning everything from little known ancient battles to the latest pharmaceutical drugs 
\cite{stats} \cite{drugs}.
All this information does not lie dormant. 
Wikipedia is the sixth most visited
site in the world, surpassing 18 billion page views in a single month. 
The most widely used general reference, Wikipedia has amassed an awe-inspiring collection of human knowledge.

It's not surprising then to see Wikipedia as the object of many studies. 
Researchers have examined the cultural dynamics among editors,
\cite{editors}
the accuracy of the content relative to traditional encyclopedias,
\cite{accuracy1}
\cite{accuracy2}
the disciplines and topics covered 
\cite{coverage}
and bias against portions of the population
\cite{bias_women}.
Wikipedia's content has also proven a powerful tool. 
Researchers have used Wikipedia to identify missing dictionary entries
\cite{missing_entries},
cluster short text
\cite{clustering},
compute semantic relatedness
\cite{semantic_relatedness},
and disambiguate meaning \cite{disambiguating}.

While these many studies have dissected and fruitfully applied Wikipedia's content,
few have examined the connections among the many articles.
Beyond information, Wikipedia is a web 
relating ideas. 
Within the body text of Wikipedia articles hyperlinks reference
other Wikipedia articles.
A hyperlink from one Wikipedia article to another is a way to relate the ideas
conveyed in the two articles 
\cite{relevance}.
The notion that hyperlinks convey information about the content of a
page has proved successful in numerous domains from search engine algorithms 
such as PageRank 
\cite{pagerank} 
to topic classification
\cite{classifier}.
We treat a hyperlink as a mechanism to connect two ideas.

The authors of a Wikipedia article choose where and whether to include a 
hyperlink reference to another Wikipedia article.
Authors must include hyperlinks in the article's markup.
In describing a particular topic, each set of authors decides which articles are the most
relevant references.
For example, the hyperlinks in the "Train" article depend only on what its authors 
decide are pertinent articles to reference.
The collection of hyperlinks emerges from the independent choice of each article's authors. 
The first hyperlink to another Wikipedia article marks the earliest moment in the body text where the authors
choose to connect the information discussed to that of another article.
By following the first hyperlink to another article in the English edition of
Wikipedia, we connect the ideas in one article to those in another, ultimately forming a directed network: 
{\it Wikipedia's First Link Network} (FLN).

\begin{figure*}[tp!]
  \centering	
  \includegraphics[width=\textwidth]{graphics/Train.pdf}  
  \caption{
    \textbf{First Link Path For "Train"}
    We follow the first link to another Wikipedia article
    in the main body of the article---the area
    inside the green rectangle, which excludes 
    side bar elements, the navigation bar and title. 
    In this example, the first link to another Wikipedia article is "Rail Transport." 
    We can again select the first link on the "Rail Transport"
    article, repeating the process to 
    form a path of First Links.
  }
  \label{fig:Train First Links}
\end{figure*}

What information could the first links possibly reveal?
Inspired by the claim that the majority of first links lead to the 
"Philosophy" article---popularized by an xkcd comic and subsequently
discussed in blog posts ---we holistically study 
Wikipedia's FLN as a map relating areas of human knowledge. 
The map is a hierarchical taxonomy where "Train" links to a parent node, 
"Rail Transport," while many child nodes, "Steel" and "Horsepower," link up to 
"Train." Unlike previous taxonomies
created by individuals
\cite{locke}
\cite{descartes}
\cite{aristotle} 
or a select group of individuals 
\cite{hist_thesaurus}, 
the organization of ideas in the FLN 
does not emerge from a concerted centralized effort, 
but rather from the independent choice of every article's authors.
The FLN is a wealth of relations among all the inventions, places,
figures, objects, and events described in Wikipedia, across space and time.

Our goal is to gain insight into how the information on Wikipedia is organized and related
by studying the structure of the FLN.
In a novel approach, we develop metrics to capture 
the dominant features of the FLN's structure.
We measure dynamics of the FLN as a flow, quantifying 
aspects of the FLN from the accumulation of first links around articles 
to the influence an article exerts in shaping the FLN.
Together with cycles, n-degree, depth, and the content of the articles, 
we build our analysis of the relations among the ideas in Wikipedia.

\section*{Traversing the First Link Network}

To understand the structure of Wikipedia's First Link Network, we
aimed to characterize the dynamics of the flow from one article to another. 
Do links tend towards a particular article, group of articles, or groups of articles? 
What is the flow of links through a typical article?
What types of cycle (loop) structures exist in the network?
What are typical paths from one article to loop or an invalid link? 
Are there exceptionally long or short paths? 
Are there articles funneling the flow of links towards a particular path?
In answering these questions about the directed network, we aimed to characterize 
how so many independent articles relate to each other.

The n-degree, or number of links directly pointing to a particular article,
inadequately captures the dynamics needed to answer these questions.
The n-degree measures only the particular set of links to a particular article, 
rather than the richer dynamics of how the links flow through the network: 
where articles tend, what the typical and atypical 
paths through the network are, which articles funnel relatively more links and so on.

Consequently, to capture the dynamics of how the First Links flow, we actually traverse every possible path through the FLN. 
The algorithm for traversing the network begins by selecting any article. 
We then proceed to the next article by following the First Link---recalling a First Link is a link in the main body of the article leading to another Wikipedia article. 
We repeat until the First Link is invalid or repeated to form a path. 
The collection of articles is path-connected conveying a flow of concepts from one article to another. 
The method is order agnostic with respect to which articles are selected first.
As long as each article is selected eventually, the resulting metrics are equivalent.

We developed three metrics to uncover structure: 
traversal visits, traversal funnels, and path length. 
Each metric captures one essence of how the links flow. 
Traversal visits gauge the accumulation of links; 
traversal funnels gauge the influence on link path; 
path length measures the number of First Links to a repeated or invalid link.

\begin{figure*}[tp!]
  \centering	
  \includegraphics[width=\textwidth]{graphics/traversal_visit_algo_figure.pdf}  
  \caption{
    \textbf{Traversal Visit Algorithm on a sample network}
     The traversal visit vectors are an adjacency matrix for the paths through the network: the first column is indicates the path formed starting with article A. The number of traversal visits for article A is then the number of paths containing A or the sum of the first row in our matrix:
     $\sum_{i=1}^7 A_{\text{visit, i}} = 7$.
  }
  \label{fig:Traversal Visits}
\end{figure*}

The first metric, {\it traversal} visits, gauges where the flow of first accumulates by associating a count to each article. 
Every path through the article increments the associated count by one. We can construct a vector to represent the articles incremented on a particular path, combining these vectors to form a matrix. 
The total accumulated counts for a given article is then simply the sum of the row in our matrix corresponding to the article of interest. 
Node A in our example (see figure ~\ref{fig:Traversal Visits}) has 7 traversal visits, since every path includes article A. 
Whereas, node E only has two traversal visits since only two paths contain article E for example.
What we have then is a metric signaling which article contains a larger accumulation of first links.

While traversal visits measure accumulation, we cannot gauge whether a particular article exerts a greater influence on flow of the first link paths. 
To distinguish between an article that simply happened to fall within a cycle from an article actually funneling the flow of first links, we developed a second metric called traversal funnels. 
We traverse the network in the same manner, but end our paths once we enter a cycle. 
We again increment the associated count of each article by one if the path up to a cycle contains the article. 
We call the accumulated count the number of {\it traversal funnels}. By measuring traversal funnels, we quantify
which articles funnel more link paths in a particular direction or cycle 
(see figure ~\ref{fig:Traversal Funnels})
.

\begin{figure*}[tp!]
  \centering	
  \includegraphics[width=\columnwidth]{graphics/traversal_funnel_algo_figure.pdf}
  \caption{
    \textbf{Traversal Funnel Algorithm on a sample network}
  The algorithm for traversal funnels is identical to the previous algorithm for traversal visits with one alteration: the path ends at the start of a cycle to distinguish articles directing a path into a cycle from articles that simply happen to be in a highly traversed path. We can construct similar vectors by considering each path through the network, measuring traversal funnels for a particular article as the sum of the entries in its corresponding row. For example
  the number of traversal funnels for article $E$ is 
  $\sum_{i=1}^7 E_{\text{funnel, i}} = 2$.}
  \label{fig:Traversal Funnels}

\end{figure*}

Finally, we track the number of links until a repeated article or invalid link, to find the typical path length within the network.
By recording the history of articles we traversed, we are also able to uncover several other network characteristics including basins, defined as a collection of path-connected articles ranked by traversal visits, and cycles of various lengths.
The three metrics: traversal visits, traversals funnels, and path length, along with our path history
yield a powerful arsenal of information with which to study the network.

\section{Discoveries}

@@

\begin{figure*}[tp!]
  \centering	
  \includegraphics[width=\columnwidth]{graphics/traversals_per_article.png} 
  \caption{
    \textbf{Distribution of Traversal Visits}
  }
  We fit the distribution to a linear log-log model by considering the log (base 10) transformed rank of each article against log (base 10) transformed the number of traversal visits. 
  The model explains $86\%$ of the variation in the data, yielding an r-value of $-0.93$ 
  and a power law exponent of $-0.636$. The horizontal flattening around the highest
  ranking articles is a result of the cyclic structure (see discussion on cycles).
  \label{fig:Distribution of Visits}

\end{figure*}

To more accurately gauge the distribution, we construct a log-log plot of the entire dataset: log(traversal visits) against log(rank). We observe a fairly linear fit, as is characteristic of scale-free networks, with an r-value of -.93. A handful of the highest ranking articles contain a disproportionate number of traversal visits, while most have none. The skew in the distribution is not terribly surprising when considering the heuristic of how the links flow: from specific to general.

@@

\begin{figure*}[tp!]
  \centering	
  \includegraphics[width=\columnwidth]{graphics/articles_ranked.png}
  \caption{
    \textbf{highest ranking articles by number of traversal visits}
  }
  We compute the number of traversal visits for each article in the FLN (see 
  Traversal Algorithm section for details). In doing so, we can rank each article
  by the accumulation of first links. Articles with a greater number of traversal visits
  mark greater points of first link accumulation. The highest ranking articles by traversal visits reveals where the greatest accumulation occurs.
  \label{fig:highest visits}
\end{figure*}

@@

\begin{figure*}[tp!]
  \centering	
  \includegraphics[width=\textwidth]{graphics/2_cycles.pdf}
  \caption{
    \textbf{highest ranking 2-cycles}
  }
  \label{fig:2-cycles}

\end{figure*}

@@

\begin{figure*}[tp!]
  \centering	
  \includegraphics[width=\textwidth]{graphics/3_cycles.pdf}
  \caption{
    \textbf{highest ranking 3-cycles}
  }
  \label{fig:3-cycles}
\end{figure*}

@@

\subsection{Basins}

We can group articles lying on the same path to identify {\it basins}: 
a group of path-connected articles.
Since cycles identify only groups of articles with a closed set of links, 
we additionally measure and rank basins to capture groups of closely related
articles branching outside of a cycle into the rest of the FLN.
Akin to river networks, these basins are areas of FLN accumulation with a path 
flowing outwards to the rest of the FLN.
The highest ranking basins by the number of traversal visits are groups of articles
around "Philosophy" given the high number of traversal visits "Philosophy" has. 
However, restricting our attention to the 
 single articles around which basins formed,
 we find high ranking basins around similarly foundational ideas:
"Community", "Landmass", "Federal Government", "Presentation", and "Belief System". 
Around each article is a sets of related ideas. Around "community" are groups of articles expressing similar ideas: "Political Union," "State," and "Federation." 
These concepts naturally emerge from the First Link Network potentially indicating pillars, which 
anchor specific knowledge in a broader, simpler concept.

\begin{figure*}[tp!]
  \centering	
  \includegraphics[width=\columnwidth]{graphics/basins.pdf}
  \caption{
    \textbf{basins around a single article}
  }
  \label{fig:basins}

\end{figure*}

@@

\begin{figure}[tp!]
  \centering	
  \includegraphics[width=\columnwidth]{graphics/path_lengths_boxplot.png}
  \caption{
    \textbf{Path Length Distribution}
  }
  \label{fig:Path Length Distribution}
\end{figure}

\begin{figure*}[tp!]
  \centering	
  \includegraphics[width=\columnwidth]{graphics/top_1k_path_length.png}  
  \caption{
    \textbf{Path Length Distribution for Top 1k articles}
  }
  \label{fig:Top 1k Path Length Distribution}
\end{figure*}

@@

Measuring traversal funnels reveals a dramatically different structure where "Philosophy" stands unmatched by orders of magnitude:
\begin{figure*}[tp!]
  \centering	
  \includegraphics[width=\textwidth]{graphics/funnels.pdf}
  \caption{
    \textbf{Funnels}
  }
  \label{fig:Funnels}
\end{figure*}

@@

%========================Conclusion====================================

\section{Reflections}

@@

\acknowledgments
Thank you to my friend RJ for pointing out the xkcd comic describing how the majority of links lead to "Philosophy," inspiring this research.

@@

\section*{Appendix}

\subsection*{Constructing The First Link Network}

Wikipedia renders and stores articles in MediaWiki markup, a markup language with syntax and keywords to format and mark elements in a page. Along with special syntax for links, MediaWiki markup includes templates for audio files, images, and side-bar
information.
While a human could accurately identify the First Link, to map the entire First Link Network of 11 million articles, we needed to programmatically untangled the body text from side-bar, header box, and invalid link elements.

While some libraries exist for MediaWiki Markup,
Approaches using existing libraries led to several bugs 
including trouble with nested links, nested parenthesis, unclosed tags, escape characters 
as well as compatibility with other libraries used to parse the XML.
Consequently,  we developed an algorithm for parsing the First Link in the XML version of each article.
Our parsing algorithm aimed to: 
1) accurately identify the First Link among other page elements, and 
2) efficiently do so---that is without needing for several passes through the data.
To process an article in a single pass, we developed a hierarchical system of flags:

\begin{figure*}[tp!]
  \centering	
  \includegraphics[width=\columnwidth]{graphics/flags.pdf}  
  \caption{
    \textbf{Parsing Algorithm of Wikipedia's XML dump}
     The highest flag in the hierarchy indicates a Wikimedia template used to mark an element in the side bar, display an image, link to an external file, or another Wikimedia project outside of Wikipeida. Next, to catch any remaining elements outside the main body we have a second flag for <ref>, <div> elements. Finally, we identify parenthesis to ensure we do not capture a link to a pronunciation key.
  }
  \label{fig:parsing algorithm}
\end{figure*}

To process the entirety of Wikipedia, we distributed the parsing and processing of the XML dump
across 112 cores of the UVM supercomputer cluster
\cite{vacc}
We then joined the results to form a hash table containing every Wikipedia article and its corresponding
First Link. The resulting network map is the basis of our analysis.

\subsection*{Supplementary Tables}

\begin{figure*}[tp!]
  \centering	
  \includegraphics[width=\columnwidth]{graphics/top_2loops.png}
  \caption{
    \textbf{highest ranking 2-Cycles}
  }
  \label{fig:Highest Ranking 2-Cyles}
\end{figure*}
